---
output: html_document
---
## ANOVA Analysis of Infant Rat Weight Gain
### From "Plane Answers to Complex Questions" by Christensen, Table 7.4

```{r, echo=FALSE, include=FALSE}
source("C:\\Users\\Todd\\R\\StatisticsProjects\\TwoWayAnova.R")
```

  
### Overview
This summarizes an ANOVA analysis of the effect of litter genotype and foster
mother genotype on infant rat weight gain. ANOVA calculations are done
using native R functions, and duplicated using custom functions that
implement concepts from linear model theory.

  
### Summary of Results
The ANOVA results show strong evidence of an effect on infant rat weight
gain due to foster mother genotype (p-value `r round(outNoI[2, 5], 3)`).
A post-hoc analysis of the foster mother genotypes shows the following:

* There is strong evidence that weight gain for genotype MF is greater than
  weight gain for genotype MJ (p-value `r tOutB[5, 4]`). The mean difference
  in weight gain between genotypes MF and MJ is
  `r abs(tOutB[5, 1])` with `r 100*tukeyLevel`%
  CI (`r abs(tOutB[5, 3])`, `r abs(tOutB[5, 2])`).

* There is suggestive evidence that weight gain for genotype MA is greater than
  weight gain for genotype MJ (p-value `r tOutB[3, 4]`). The mean difference in
  weight gain between genotypes MA and MJ is `r abs(tOutB[3, 1])` with
  `r 100*tukeyLevel`% CI (`r abs(tOutB[3, 3])`, `r abs(tOutB[3, 2])`).

  
### Two-Way ANOVA with Interaction
Here is a Type I ANOVA analysis that includes the Litter:Mother interaction:

```{r echo=FALSE}
outWithI
```

The large p-value for the Litter:Mother interaction shows no evidence of an
interaction effect.

It is interesting to compare the ANOVA results with a
plot of the interaction between the Litter and Mother effects:

```{r echo=FALSE}
prettyInteractionPlot(dfRats, 1, 2, 3, "Litter:Mother Interaction")
```

Although the Mother MA and MF genotypes seem to behave differently as the Litter genotype
changes, the other two Mother genotypes show little interaction with the Litter genotypes.
So this plot seems consistent with the p-value for the interaction effect. Since
there is no evidence of a significant interaction effect, we will drop the interaction
effect from the model and do a two-way ANOVA analysis without the interaction effect.
  
    
### Two-Way ANOVA with Main Effects Only
These results of a two-way ANOVA analysis without interaction test the significance of
each main effect in the presence of the other main effect:

```{r echo=FALSE}
outNoI
```

The large p-value for the Litter effect shows no evidence of an effect due to
Litter genotype. The small p-value for the Mother effect is strong evidence
of an effect due to Mother genotype.
  
    
### Assumptions of ANOVA Analysis
The ANOVA results above are dependent on three assumptions about the model
residuals: normal distribution, constant variance, and independence. Those
three assumptions were tested as shown below:

1. Check for normally distributed residuals:
    ```{r echo=FALSE}
    shapiro.test(out$residuals)
    ```
    
    The moderately small p-value suggests the residuals may not be normally
    distributed.

    The QQ plot of residuals also suggests a possible deviation from normality.
    This is especially evident at the far left and far right of the plot:
    ```{r echo=FALSE}
    prettyQQ(out$residuals, "QQ Plot of Residuals")
    ```
    
    Such a moderate deviation from normality is acceptable, especially given the
    fairly large sample size of `r dim(dfRats)[1]`.
     
2. Check for constant variance of residuals:
    ```{r echo=FALSE}
    hov(WeightGain ~ Mother, data = dfRats)
    ```
    The Brown-Forsyth test for constant variance across Mother genotypes
    yields a large p-value. This shows no evidence that the residual variances
    differs across Mother genotypes:
    
    We can also check the homogeneity of variances by looking at a scatter plot of
    residuals vs. fitted values, categorized by Mother genotype:

    ```{r echo=FALSE}
    prettyResVsFit(dfOut, 1, 2, FALSE, "Residuals vs. Fitted Values, Categorized by Mother", 4)
    ```
    
  + This scatter plot does not show a significant difference in variances
    across Mother genotypes. It also shows fairly constant variability
    across all data.
      
  + In addition, this plot does not show any systematic patterns in the residuals
    that would indicate the model is missing some significant effect.
        
3. Check for independence among residuals:  
   Not enough is known about the circumstances of the experiment to evaluate
   this assumption.
  
    
### Pair-Wise Differences Among Mother Genotypes
Now that we see evidence of a significant difference among Mother genotypes,
information about the pair-wise differences is interesting. We examined
those differences two ways: a categorized box plot, and the Tukey HSD test:

1. A box plot categorized by Mother genotype indicates the biggest difference
   is between genotypes MJ and MF:
   ```{r echo=FALSE}
   prettyBoxPlot(dfRats, 1, 3, "Weight Gain Categorized by Mother")
   ```
    
2. Tukey HSD test:
   ```{r echo=FALSE}
   kable(tOutA$Mother)
   ```
       
  + The small p-value for the difference between genotypes MJ and MF is strong
    evidence of a difference between those two genotypes. This is consistent with
    the categorized box plot above.
      
  + The moderately-sized p-value for the difference between genotypes MJ and MA
    is suggestive of a difference between those two genotypes. This is also
    consistent with the categorized box plot above.

    
### Outliers
Sometimes ANOVA outliers contain interesting information. A scatter plot of
studentized residuals vs. fitted values shows no outliers.
  ```{r echo=FALSE}
  prettyResVsFit(dfOut, 1, 3, TRUE, "Studentized Residuals vs. Fitted Values")  
  ```

  

